A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-30 code/figure2.Rmd
update text and figures

```{r figure-2-plot, out.height='90%', out.width='90%',fig.cap="Figure 2: Pan-cancer enrichment of pathway disruptions. Waterfall plot; columns represent tumors. Each row represents a pathway. The pathways are ranked such that the pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the top pathway disruptions successively. All 377 pathways are shown. Red signifies pathway disrupted pathways, grey signifies not disrupted."}

@@

>>>>

```{r figure-2-plot, out.height='90%', out.width='90%'}

@@

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 2: Pan-cancer enrichment of pathway disruptions.** a) Waterfall plot; columns represent tumors. Each row represents a pathway. The pathways are ranked such that the pathway with the most disruptions across all 7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the top pathway disruptions successively. The top 100 most mutated pathways are shown. Red signifies pathway disrupted pathways, grey signifies not disrupted. b) Heatmap shows the relative enrichment of each pathway (columns) within each numbered cluster (rows). Effect size is displayed as colors (orange, enrichment; purple, depletion). c) Proportion of mutated samples in the each significant pathway (columns; union set of pathways with effect size >= 0.30 in each cluster) within each numbered cluster (rows).</p>

### 2020-07-30 text/manuscript.Rmd
update text and figures

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(foreach)
library(sqldf)
library(tidyverse)
library(ggthemes)
library(data.table)
library(biomaRt)
library(dplyr)
library(GenomicRanges)
library(SummarizedExperiment)
library(reactome.db)
library(plyr)
library(cluster)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(dbscan)
library(class)
library(TCGAbiolinks)
library(patchwork)
library(ggplotify)
library(ComplexHeatmap)
library(knitr)
library(kableExtra)
library(UpSetR)
library(cowplot)
library(doRNG)
library(uwot)
library(janitor)
library(FactoMineR)
library(seriation)
```

@@

To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data because were most appropriate for our
study given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to start with somatic mutations in exome
sequencing data because the affected target gene is known unambiguously.
Therefore, we focused on all missense, nonsense, frameshift, and splicing
mutations. In order to minimize bias from well-studied diseases and processes,
we selected 377 Reactome pathways [reactome.org](https://reactome.org/) of
interest corresponding to basic cellular processes and biochemical pathways,
excluding gene sets that correspond to miscellaneous categories ( _e.g._
“transcription factors”) or disease associations ( _e.g._ “mutated in colon
cancer”) and filtered our gene list on membership in these pathways (<mark>total
of x genes;</mark> see methods for details).

It does not matter for our purposes whether a single tumor is “enriched” for
mutations of individual pathways, chiefly because such metrics favor larger
pathways ( _i.e._ pathways with more member genes) over smaller ones. Data
binarized by pathway are likely to be noisy for several reasons. First, point
mutations can be deleterious (attenuating or hypomorphic) or activating
(neomorphic) in genes that act functionally as either activators or repressors.
Within the scope of this study we accept this uncertainty and assume mutations
are generically disruptive to the pathway regardless of whether they are
activating or deactivating or whether they affect a gene that we might classify
as one thing or another ( _e.g._ "Tumor suppressor" or "Oncogene"). Second, we
know that low-expressed genes and non-expressed genes are known to accumulate
mutations at a higher rate due to transcription coupled repair
[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
issue, we identified genes with low expression in each type of cancer and
eliminated them for that cancer type only (see methods for details).
<strike>Highly expressed genes could also have high mutation rates owing to
transcription induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt
that this mechanism results in cell-type-specific biases that might be
biologically meaningful for predisposition to different classes of cancer
however, and therefore chose not to exclude these genes from our analysis. After
selecting our pathways and genes, we then compiled a matrix of the pathways,
assigning a Boolean value (a 1 or 0, indicating whether any pathway member gene
carries a mutation) to the pathway regardless of the number of genes carrying
mutations (**figure 1A**).

```{r child='../code/figure1.Rmd'}
```
In order to classify tumors using this dataset, we used multiple correspondence
analysis (MCA) [@JSSv025i01]. First we determined the number of dimensions
containing useful information (**figures 1B, C**). Then we performed a UMAP
analysis, both in order to summarize the MCA graphically, and as a preprocessing
step to boost the performance of density based clustering. The resulting map was
notable for its lobed structure, with several large, robust projections
reproducible regardless of random seed setting. A representative version of this
3D UMAP is shown in **figure 1D** as a 2D projection, rotated to enhance the
visibility of the major features. Following this spatial mapping we attempted to
define groupings of similar tumors within the spatial map using HDBSCAN, which
performs hierarchical clustering and provides metrics of cluster stability and
probabilities of cluster membership for each node. However, HDBSCAN is sensitive
to several parameters; key for our analysis are the minimum number of tumor
samples in a cluster that capture the maximum number of tumor samples, measured
by probability of membership of $\geq 5\%$ in at least one class. Thus, we
created a score metric as the fraction of classified tumors with max probability
$< 5\%$ in one class and chose a cluster size of 95 to minimize the score
function (**figure 1E**). HDBSCAN with these settings resulted in 12 distinct
high-density clusters which we then projected onto the UMAP (**figure 1F**).
This classified <mark>x out of 7,607 tumors</mark> but still resulted in a
significant fraction of unclassified tumor samples. Since we ultimately wish to
be able to classify any tumor using this scheme, we performed k Nearest
Neighbors (kNN) analysis, which computes a similarity metric to every tumor in
the set and then lets the $k$ most similar tumors "vote" as to the identity of
the query tumor sample based on their classification labels. We set $k$ to be
the square root of the number of tumor samples (87). Using this method, we
assigned cluster membership to the remaining tumors (**figure 1G**).

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types (with some notable
exceptions) were not heavily biased in one class, and all well-represented
cancer types had representative tumors in every class (for example **figure 1I**
and full tumor profiles in **supplementary figure S1**, see also interactive
media from **supplementary file 1 (html)**), suggesting that, in principle at
least, our novel pathway-disruption classification method identifies molecular
pathologies largely independent of tissue-of-origin considerations. As an
example of one type of cancer that does have a biased pathway profile,
pancreatic adenocarcinoma (PAAD) was biased in class 8 & 9, which is
distinguished  by KRAS mutations among others (**supplementary figure S3**). But even PAAD comprises tumors
from <mark>check; 7 out of 10 of the remaining classes</mark>, meaning that
patients hosting these tumors have potentially different underlying molecular
pathologies.

@@

Many cancers have molecular or histological subtypes defined based on gene
expression or other -omics profiles or pathology lab results. These subtypes
often have different standards of care owing to different overall drug
sensitivity (or other factors). If the histological subtypes represent true
molecular phenotypes, one predicts that histological subtypes should segregate
with our pathway-based classes, therefore providing support for the classes as
proxies for molecular pathology sub-typing. To our surprise, we find a similar
result to the previous analysis of cancer types projected onto the UMAP of
pathway disruptions. For example we projected annotations for each of the breast
cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
normal-like, and lumenal A and B subtypes onto the UMAP. These are among the
most heavily studied molecular subtypes in cancer, which each have different
prognoses and standards of care. We did not observe any exclusive segregation by
pathway for these subtype annotations (**Figure 1J** and **supplemental
file 1**). We also projected histological subtype data for the remaining cancers
(see **supplementary figure S2** for the full set of projections); we find that
the subtypes, though often biased towards one or more classes, are almost never
exclusive to one of the classes. We interpret these data in aggregate to mean
that our pathway disruption classes do not correspond to molecular subtypes
within the parent cancer type. Furthermore, this result raises the possibility
that, molecular and histo-pathological subtypes could represent different
precursor cell origins within each cancer type. This interpretation still leaves
open the possibility that different precursor cell types, being susceptible to
different teratogens and selection pressures ( _e.g._ insulin dependence) could
be biased toward one or more of the pathway disruption profiles.

@@

There are several well-known familial cancer-causing mutations that have been
interrogated extensively for differences in basic biology, survival and
treatment outcomes. However, the functions of these genes are related to risk
factors such as genome stability, proof-reading and DNA damage repair, and
telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
breast, prostate and ovarian cancers. The mechanism of risk is thought to
involve loss of heterozygosity [@brca-loh], so we projected the somatic
mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
segregation of these mutations into specific clusters (**figure 1K**). We made
similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH_,
_PMS1_ and _PMS2_; _BRIP1_, _RAD15_, _CHEK2_ and _APC_. None of these genes
except for _APC_ exhibited any remarkable specificity with respect to cluster
assignment (**supplementary figure S3**). To look at other risk factors such as
maintenance of DNA methylation levels and telomere length we projected somatic
mutations of the _TET2_ and _TET3_ genes, plus telomerases 1, 2, and 3, and
observed similar lack of segregation by cluster (**supplementary figure S3**).

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(foreach)
library(sqldf)
library(tidyverse)
library(ggthemes)
library(data.table)
library(biomaRt)
library(dplyr)
library(GenomicRanges)
library(SummarizedExperiment)
library(reactome.db)
library(plyr)
library(cluster)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(dbscan)
library(class)
library(TCGAbiolinks)
library(patchwork)
library(ggplotify)
library(ComplexHeatmap)
library(knitr)
library(kableExtra)
library(UpSetR)
library(cowplot)
library(doRNG)
library(uwot)
library(janitor)
library(FactoMineR)
library(seriation)
library(rtracklayer)
library(poibin)
```

@@

To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data because were most appropriate for our
study given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to start with somatic mutations in exome
sequencing data because the affected target gene is known unambiguously.
Therefore, we focused on all missense, nonsense, frameshift, stop-gain, untranslated
region, and splicing mutations. In order to minimize bias from well-studied diseases
and processes, we selected 377 Reactome pathways [reactome.org](https://reactome.org/) 
of interest corresponding to basic cellular processes and biochemical pathways,
excluding gene sets that correspond to miscellaneous categories ( _e.g._
“transcription factors”) or disease associations ( _e.g._ “mutated in colon
cancer”) and filtered our gene list on membership in these pathways (total
of 9040 genes; see methods for details).

It does not matter for our purposes whether a single tumor is “enriched” for
mutations of individual pathways, chiefly because such metrics favor larger
pathways ( _i.e._ pathways with more member genes) over smaller ones. Data
binarized by pathway are likely to be noisy for several reasons. First, point
mutations can be deleterious (attenuating or hypomorphic) or activating
(neomorphic or hypermorphic) in genes that act functionally as either activators or repressors.
Within the scope of this study we accept this uncertainty and assume mutations
are generically disruptive to the pathway regardless of whether they are
activating or deactivating or whether they affect a gene that we might classify
as one thing or another ( _e.g._ "Tumor suppressor" or "Oncogene"). Second, we
know that low-expressed genes and non-expressed genes are known to accumulate
mutations at a higher rate due to transcription coupled repair
[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
issue, we identified genes with low expression in each type of cancer and
eliminated them for that cancer type only (see methods for details). 
<strike> Highly expressed genes could also have high mutation rates owing to
transcription induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt
that this mechanism results in cell-type-specific biases that might be
biologically meaningful for predisposition to different classes of cancer
however, and therefore chose not to exclude these genes from our analysis. After
selecting our pathways and genes, we then compiled a matrix of the pathways,
assigning a Boolean value (a 1 or 0, indicating whether any pathway member gene
carries a mutation) to the pathway regardless of the number of genes carrying
mutations (**figure 1A**). </strike>

```{r child='../code/figure1.Rmd'}
```
In order to classify tumors using this dataset, we used multiple correspondence
analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
containing useful information. (**figures 1B, C**). Then we performed a UMAP
analysis, both in order to summarize the MCA graphically, and as a preprocessing
step to boost the performance of density based clustering. The resulting map was
notable for its lobed structure, with several large, robust projections
reproducible regardless of random seed setting. A representative version of this
3D UMAP is shown in **figure 1D** as a 2D projection, rotated to enhance the
visibility of the major features. Following this spatial mapping we attempted to
define groupings of similar tumors within the spatial map using HDBSCAN, which
performs hierarchical clustering and provides metrics of cluster stability and
probabilities of cluster membership for each node. However, HDBSCAN is sensitive
to several parameters; key for our analysis are the minimum number of tumor
samples in a cluster that capture the maximum number of tumor samples, measured
by probability of membership of $\geq 5\%$ in at least one class. Thus, we
created a score metric as the fraction of classified tumors with max probability
$< 5\%$ in one class and chose a cluster size of 92 to minimize the score
function (**figure 1E**). HDBSCAN with these settings resulted in ten distinct
high-density clusters which we then projected onto the UMAP (**figure 1F**).
This classified 6,038 out of 7,607 tumors but still resulted in a
significant fraction of unclassified tumor samples. Since we ultimately wish to
be able to classify any tumor using this scheme, we performed k Nearest
Neighbors (kNN) analysis, which computes a similarity metric to every tumor in
the set and then lets the $k$ most similar tumors "vote" as to the identity of
the query tumor sample based on their classification labels. We set $k$ to be
the square root of the number of tumor samples (87). Using this method, we
assigned cluster membership to the remaining tumors (**figure 1G**).

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types (with some notable
exceptions) were not heavily biased in one class, and all well-represented
cancer types had representative tumors in every class (for example **figure 1I**
and full tumor profiles in **supplementary figure S1**, see also interactive
media from **supplementary file 1 (html)**), suggesting that, in principle at
least, our novel pathway-disruption classification method identifies molecular
pathologies largely independent of tissue-of-origin considerations. As an
example of one type of cancer that does have a biased pathway profile,
pancreatic adenocarcinoma (PAAD) was biased in class 8, which is
distinguished  by KRAS mutations among others (**supplementary figure S3**). 
But even PAAD comprises tumors from the nine remaining classes, meaning that 
patients hosting these tumors have potentially different underlying molecular
pathologies.

@@

Many cancers have molecular or histological subtypes defined based on gene
expression or other -omics profiles or pathology lab results. These subtypes
often have different standards of care owing to different overall drug
sensitivity (or other factors). If the histological subtypes represent true
molecular phenotypes, one predicts that histological subtypes should segregate
with our pathway-based classes, therefore providing support for the classes as
proxies for molecular pathology sub-typing. To our surprise, we find a similar
result to the previous analysis of cancer types projected onto the UMAP of
pathway disruptions. For example we projected annotations for each of the breast
cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
normal-like, and luminal A and B subtypes onto the UMAP. These are among the
most heavily studied molecular subtypes in cancer, which each have different
prognoses and standards of care. We did not observe any exclusive segregation by
pathway for these subtype annotations (**Figure 1J** and **supplemental
file 1**). We also projected histological subtype data for the remaining cancers
(see **supplementary figure S2** for the full set of projections); we find that
the subtypes, though often biased towards one or more classes, are almost never
exclusive to one of the classes. We interpret these data in aggregate to mean
that our pathway disruption classes do not correspond to molecular subtypes
within the parent cancer type. Furthermore, this result raises the possibility
that, molecular and histo-pathological subtypes could represent different
precursor cell origins within each cancer type. This interpretation still leaves
open the possibility that different precursor cell types, being susceptible to
different teratogens and selection pressures ( _e.g._ insulin dependence) could
be biased toward one or more of the pathway disruption profiles.

@@

There are several well-known familial cancer-causing mutations that have been
interrogated extensively for differences in basic biology, survival and
treatment outcomes. However, the functions of these genes are related to risk
factors such as genome stability, proof-reading and DNA damage repair, and
telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
break repair [@moynihan-1999-brca1;@davies-2001-brca2] and confer risk for
breast, prostate and ovarian cancers. The mechanism of risk is thought to
involve loss of heterozygosity [@brca-loh], so we projected the somatic
mutations for _BRCA1_ and _-2_ genes onto the UMAP, but did not observe
segregation of these mutations into specific clusters (**figure 1K**). We made
similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_,
_MLH3_, _PMS1_ and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes
except for _APC_ exhibited any remarkable specificity with respect to cluster
assignment (**supplementary figure S3**). To look at other risk factors such as
maintenance of DNA methylation levels and telomere length, we projected somatic
mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and
observed similar lack of segregation by cluster (**supplementary figure S3**).
